---
title: "AQ_curves_example"
author: "Nick Tomeo"
date: "11/7/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction
This document is meant as an introduction to the AQ_curve functions I put together. My hope is that some examples of how they work will serve as a quick instructional how-to.

### Data

To begin with let's load and familiarize with some of the example data files.
```{r packages, include = FALSE}
if (!require("RCurl", quietly=TRUE)) {
    install.packages("RCurl")
    library("RCurl")
}
```

```{r load_data}
a_thal = read.csv(text = getURL(
      "https://raw.githubusercontent.com/Tomeopaste/AQ_curves/master/ExampleDataFiles/athal_LRCs.csv"),
      header = TRUE)
dwarf = read.csv(text = getURL(
      "https://raw.githubusercontent.com/Tomeopaste/AQ_curves/master/ExampleDataFiles/dwarfSunflower.csv"),
      header = TRUE)

# Create a column with a unique identifier for each curve in the data frame
dwarf$ids = paste0(dwarf$Light, "_", dwarf$Replicate)
head(a_thal)
head(dwarf)
```

Now we have `r length(unique(a_thal$ids))` light response curves measured on an assortment of *Arabidopsis thaliana* ecotypes. And, `r length(unique(dwarf$ids))` curves measured on dwarf sunflowers, 1-each at each combination of nutrientsXirradiance (high & low) treatments. These look like your typical curves:

```{r basic_curve_plots}
par(mfrow=c(1,1), oma=c(1,1,1,1), mar=c(4,4,1,1))
plot.ids = unique(dwarf$ids)
for (i in seq_along(plot.ids)) {
      plot(dwarf[dwarf$ids == plot.ids[i], "Photo"] ~ 
                 dwarf[dwarf$ids == plot.ids[i], "PARi"],
           xlab = "PARi", ylab = "Photo", pch=i,
           xlim=c(-2,1600), ylim=c(
                 min(dwarf$Photo)-2, max(dwarf$Photo)+2
           ))
      if(i < 3.5){par(new=TRUE)}
}
```

### Fitting function

The `fit_AQ_curve()` function in the script https://github.com/Tomeopaste/AQ_curves/blob/master/AQ_curve_function.R fits the light response data to a non-rectangular hyperbola model (Equation 6 in Lobo et al. 2013, Photosynthetica, v51, doi.org/10.1007/s11099-013-0045-y):

$$A_{N} = \frac{\Phi*\mathit{Q}*A_{sat}-\sqrt{(\Phi*\mathit{Q}*A_{sat})^{2}-4*\Theta*\Phi*\mathit{Q}*A_{sat}}}{2*\Theta }-R_{d}$$
where, A~N~ is the net photosynthetic rate, $\Phi$ is quantum efficiency, *Q* is irradiance in the PAR range, A~sat~ is the light-saturated photosynthetic rate, $\Theta$ is a convexivity parameter relating the curvature of the response, and R~d~ is the respiration rate in the day/light.

Let's load in `fit_AQ_curve()`. I recommend clicking the little triangle next to the top line in the next code chunk (`fit_AQ_curve <- function...` ) to collapse the whole thing once you run it. That way you do not have to scroll through all 170 lines of it.
```{r fitting_function, echo = FALSE}
fit_AQ_curve <- function(df, group_id, Photo, PARi, fit_type = "onls"){
      AQ_curve_fits <- data.frame(ID = character(),
                                  Asat = numeric(),
                                  Phi = numeric(),
                                  Rd = numeric(),
                                  theta = numeric(),
                                  resid_SSs = numeric(),
                                  LCP = numeric(),
                                  Q_sat_75 = numeric(),
                                  Q_sat_85 = numeric(),  
                                  stringsAsFactors = FALSE
      )
      if(fit_type == "onls"){
      if(require("onls")){
            print("onls is loaded correctly")
      } else {
            print("trying to install onls")
            install.packages("onls")
            if(require(onls)){
                  print("onls installed and loaded")
            } else {
                  stop("could not install onls")
            }
      }
      library("onls")      
      for(i in seq_along(unique(df[[group_id]]))){
            tryCatch({
                  AQ_curve_fits[i, 1] <- unique(df[[group_id]])[i]
                  # Subset by group_ID iteratively:
                  single_curve1 <- df[df[[group_id]] == unique(df[[group_id]])[i],]
                  single_curve1$assim <- single_curve1[[Photo]]
                  single_curve1$PAR <- single_curve1[[PARi]]
                  single_curve = single_curve1[order(single_curve1$PAR),]
                  phi.as.slope <- with(single_curve,
                                       as.numeric(coef(lm(
                                             assim[1:5] ~ PAR[1:5]))[2]))
                  # Fit the curve:
                  temp.fit <- with(single_curve, # use the subset of a single curve
                                   onls(assim ~ ((Phi * PAR + Asat - 
                                                       sqrt((Phi * PAR + Asat)^2 - 
                                                                  4 * Phi * theta * 
                                                                  Asat * PAR ))
                                   )/(2*theta) - Rd,
                                   start=list(
                                         Asat = (max(assim)),
                                         Phi = phi.as.slope,
                                         Rd = -min(assim),
                                         theta = 0.5),
                                   control = list(maxiter = 50),
                                   verbose = FALSE
                                   )
                  )
                  AQ_curve_fits[i, 2] <- as.numeric(coef(temp.fit)[1]) # asat 
                  AQ_curve_fits[i, 3] <- as.numeric(coef(temp.fit)[2]) # Phi
                  AQ_curve_fits[i, 4] <- as.numeric(coef(temp.fit)[3]) # Rd
                  AQ_curve_fits[i, 5] <- as.numeric(coef(temp.fit)[4]) # theta
                  AQ_curve_fits[i, 6] <- sum(resid(temp.fit)^2)
                  AQ_curve_fits[i, 7] <- (as.numeric(coef(temp.fit)[3]) *(
                        as.numeric(coef(temp.fit)[3]) * as.numeric(coef(temp.fit)[4]) - 
                              as.numeric(coef(temp.fit)[1]))
                  ) / (as.numeric(coef(temp.fit)[2]) * (
                        as.numeric(coef(temp.fit)[3]) - as.numeric(coef(temp.fit)[1])
                  ))
                  AQ_curve_fits[i, 8] <- (
                        (as.numeric(coef(temp.fit)[1]) * 0.75 + 
                               (as.numeric(coef(temp.fit)[3]))) * (
                                     as.numeric(coef(temp.fit)[1]) * 0.75 *
                                           as.numeric(coef(temp.fit)[4]) +
                                           as.numeric(coef(temp.fit)[3]) *
                                           as.numeric(coef(temp.fit)[4]) -
                                           as.numeric(coef(temp.fit)[1])
                               )) / (
                                     as.numeric(coef(temp.fit)[2])* (
                                           as.numeric(coef(temp.fit)[1]) * 0.75 +
                                                 as.numeric(coef(temp.fit)[3]) -
                                                 as.numeric(coef(temp.fit)[1])
                                     ))
                  
                  AQ_curve_fits[i, 9] <- (
                        (as.numeric(coef(temp.fit)[1]) * 0.85 + 
                               (as.numeric(coef(temp.fit)[3]))) * (
                                     as.numeric(coef(temp.fit)[1]) * 0.85 *
                                           as.numeric(coef(temp.fit)[4]) +
                                           as.numeric(coef(temp.fit)[3]) *
                                           as.numeric(coef(temp.fit)[4]) -
                                           as.numeric(coef(temp.fit)[1])
                               )) / (
                                     as.numeric(coef(temp.fit)[2])* (
                                           as.numeric(coef(temp.fit)[1]) * 0.85 +
                                                 as.numeric(coef(temp.fit)[3]) -
                                                 as.numeric(coef(temp.fit)[1])
                                     ))
            }, error = function(E){cat("Error: ", conditionMessage(E), "\n")})
      }
      return(AQ_curve_fits)
      } else{
      if(fit_type == "nls"){
            for(i in seq_along(unique(df[[group_id]]))){
                  tryCatch({
                  AQ_curve_fits[i, 1] <- unique(df[[group_id]])[i]
                  # Subset by group_ID iteratively:
                  single_curve1 <- df[df[[group_id]] == unique(df[[group_id]])[i],]
                  single_curve1$assim <- single_curve1[[Photo]]
                  single_curve1$PAR <- single_curve1[[PARi]]
                  single_curve = single_curve1[order(single_curve1$PAR),]
                  phi.as.slope <- with(single_curve,
                                       as.numeric(coef(lm(
                                             assim[1:5] ~ PAR[1:5]))[2]))
                  # Fit the curve:
                  temp.fit <- with(single_curve, 
                                   nls(assim ~ ((Phi * PAR + Asat - 
                                           sqrt((Phi * PAR + Asat)^2 - 
                                                4 * Phi * theta * 
                                                      Asat * PAR ))
                                               )/(2*theta) - Rd,
                                               start=list(
                                                     Asat = (max(assim)),
                                                     Phi = phi.as.slope,
                                                     Rd = -min(assim),
                                                     theta = 0.5),
                                               control = list(maxiter = 50),
                                               algorithm = "port")
                              )
                  AQ_curve_fits[i, 2] <- as.numeric(coef(temp.fit)[1]) # asat 
                  AQ_curve_fits[i, 3] <- as.numeric(coef(temp.fit)[2]) # Phi
                  AQ_curve_fits[i, 4] <- as.numeric(coef(temp.fit)[3]) # Rd
                  AQ_curve_fits[i, 5] <- as.numeric(coef(temp.fit)[4]) # theta
                  AQ_curve_fits[i, 6] <- sum(resid(temp.fit)^2)
                  AQ_curve_fits[i, 7] <- (as.numeric(coef(temp.fit)[3]) *(
                              as.numeric(coef(temp.fit)[3]) * 
                              as.numeric(coef(temp.fit)[4]) - 
                              as.numeric(coef(temp.fit)[1]))
                              ) / (as.numeric(coef(temp.fit)[2]) * (
                              as.numeric(coef(temp.fit)[3]) - 
                                    as.numeric(coef(temp.fit)[1])
                              ))
                  AQ_curve_fits[i, 8] <- (
                              (as.numeric(coef(temp.fit)[1]) * 0.75 + 
                              (as.numeric(coef(temp.fit)[3]))) * (
                              as.numeric(coef(temp.fit)[1]) * 0.75 *
                              as.numeric(coef(temp.fit)[4]) +
                              as.numeric(coef(temp.fit)[3]) *
                              as.numeric(coef(temp.fit)[4]) -
                              as.numeric(coef(temp.fit)[1])
                              )) / (
                              as.numeric(coef(temp.fit)[2])* (
                              as.numeric(coef(temp.fit)[1]) * 0.75 +
                              as.numeric(coef(temp.fit)[3]) -
                              as.numeric(coef(temp.fit)[1])
                              ))
                  AQ_curve_fits[i, 9] <- (
                              (as.numeric(coef(temp.fit)[1]) * 0.85 + 
                              (as.numeric(coef(temp.fit)[3]))) * (
                              as.numeric(coef(temp.fit)[1]) * 0.85 *
                              as.numeric(coef(temp.fit)[4]) +
                              as.numeric(coef(temp.fit)[3]) *
                              as.numeric(coef(temp.fit)[4]) -
                              as.numeric(coef(temp.fit)[1])
                              )) / (
                              as.numeric(coef(temp.fit)[2])* (
                              as.numeric(coef(temp.fit)[1]) * 0.85 +
                              as.numeric(coef(temp.fit)[3]) -
                              as.numeric(coef(temp.fit)[1])
                              ))
                        }, error = function(E){
                              cat("Error: ", conditionMessage(E), "\n")})
                  }
                  return(AQ_curve_fits)      
            } else{print("ERROR: 'fit_type' specified incorrectly.")}
      }
}

```

Now to fit our curves we need to supply the function with the data as a data_frame plus three additional required arguments:   

1. the name of the column in the data containing net photosynthetic values   
      + `Photo = "my_photo_values"`   
2. the name of the column containing irradiance values   
      + `PARi = "my_irradiance_values"`   
3. the name of the column with unique identifying information that differntiates the individual curves.   
      + `group_id = "my_unique_curve_identifiers"`   
      
________________________________________________________________________________
________________________________________________________________________________
________________________________________________________________________________
Or, presented in R Documentation-like format.  

## Fitting photosynthetic light response curves (A~Q curves)   

#### Description    

Fits a non-rectangular hyperbola model to photosynthetic light response gas exchange data and solves for the commonly used parameters: light-saturated photosynthetic rate (A~sat~), quantum efficiency ($\Phi$), mitochondrial respiration in the light (R~d~), curvature/convexivity of light saturation ($\Theta$), light compensation point (LCP), and the irradiance required to saturate photosynthesis.
    
#### Usage    
    
`fit_AQ_curve(df, group_id, Photo, PARi, fit_type = "onls")`
    
#### Arguments    

Argument      | Description
------------- | ----------------------------------------------------------------
`df`          | A data frame with your gas exchange data that contains a column with unique identifiers for each curve (e.g., "curve_1", "curve_2", etc), a column of net photosynthetic values, and a column of irradiance/Q/PAR/PPFD values.
`group_id`    | The name of the column containing curve identification values.
`Photo`       | The name of the column containing net photosynthetic values.
`PARi`        | The name of the column containing irradiance values.
`fit_type`    | The type of regression model to use for fitting. Defaults to orthogonal difference with `onls::onls()`. Can optionally be switched to `base::nls()` by explicityly setting to "nls". More detail on this below.
________________________________________________________________________________
________________________________________________________________________________
________________________________________________________________________________
    
The function will return a data_frame that contains:  

1. The group_id's  
2. Light saturated net photosynthesis (Asat)  
3. Quantum yield (Phi)  
4. Mitochondrial respiration in the light (Rd)   
5. The curvature/convexivity factor of the curve (theta)  
6. The residual sum-of-squares (resid_SSs) from the model fit  
7. The light compensation point (LCP)  
8. PARi at 75% saturation of photosynthesis (Q_sat_75) - reliable  
9. PARi at 85% saturation (Q_sat_85) - much less reliable  

________________________________________________________________________________
________________________________________________________________________________

A basic call will look like this:  
   
`fit_AQ_curve(myData, Photo = "Photo", PARi = "PARi", group_id = "group_ids")`

________________________________________________________________________________
________________________________________________________________________________
  
Try it out with the dwarf sunflower data where the Photo and PARi columns are named "Photo" and "PARi" (the 6400 output) and group_id column is named "ids":
```{r dwarf_fits}
fit_AQ_curve(dwarf, Photo = "Photo", PARi = "PARi", group_id = "ids")
```
You can *mostly* ignore the messages that come out of `fit_AQ_curve()`, with one caveat. The first time you use `fit_AQ_curve()`, it will attempt to install the `onls` package and if the installation fails you will recieve a message alerting you to that failure. If that happens you have two options: 1) install `onls` manually, or 2) run `fit_AQ_curve()` with the alternative fitting method (see *Fitting method* below).

________________________________________________________________________________
________________________________________________________________________________

We can save the output as an object so that we can do other things with it - even if those things don't make a whole lot of sense with this example data. Here I save the sunflower data fits to the object `dwarf_fits` and run a t-test to compare A~sat~ between the plants grown at high and low irradiance.
```{r extended_dwarf_example}
dwarf_fits = fit_AQ_curve(dwarf, Photo = "Photo", PARi = "PARi", group_id = "ids")
dwarf_fits$trt = substr(dwarf_fits$ID, 1, 3) # create treatment variable
t.test(dwarf_fits$Asat ~ dwarf_fits$trt)
```
   
   
We can similarly fit the larger *Arabidosis* ecotype dataset, then summarise and plot some of the results.
```{r fitting_arabidopsis_data}
# changing the column names for demonstration purposes:
names(a_thal)[3:4] = c("irradiance", "A_n")

athal_fits = fit_AQ_curve(a_thal,
                          Photo = "A_n",
                          PARi = "irradiance",
                          group_id = "ids")
summary(athal_fits[,2:9])
hist(athal_fits$Asat, main = "Not-very useful histogram")
plot(athal_fits$Asat ~ athal_fits$Rd,
     main = "Ugly scatterplot with lm() fit-line")
abline(lm(athal_fits$Asat ~ athal_fits$Rd))
```

### Fitting method

When discussing the installation of the `onls` package above I mentioned an alternative fitting method. The default behavior is to use orthogonal difference regression to fit the curve with `onls::onls()`. In my experience this method is more reliable, and it avoids the error-in-variables problem that plagues many traditional curve fitting approaches. That being said, if for whatever reason you prefer `base::nls()`, you can use it (with `algorithm = "port"`) by including the optional argument `fit_type = "nls"`" in your `fit_AQ_curve()` call. In practice - on the curves I've tested - the differences in fit values between the two methods are too small to really matter, i.e., they're only different at approximately the sixth significant figure (see just below). *But*, `onls()` does produce fits on at least some curves where `nls()` fails, hence using it as the default.

```{r alt_fit_method}
nls_dwarf_fits = fit_AQ_curve(dwarf, # Same call except...
                              Photo = "Photo",
                              PARi = "PARi",
                              group_id = "ids",
                              fit_type = "nls") # switching to fitting w/ nls()
# Compare the fits using onls() and nls():
dwarf_fits[,c(2:5,7:9)] - nls_dwarf_fits[,c(2:5,7:9)]
dwarf_fits[,c(2:5,7:9)] / nls_dwarf_fits[,c(2:5,7:9)]
```

Again, when both methods work, it is clear that the differnces are trivial.

### Assessing the quality of model fit

Producing fits, and whether or not they are good fits, are two different things. `fit_AQ_curve()` currently only calculates the residual sum-of-squares for the fit model - I am happy to add a better metric if anybody wants to convince me of such. resid_SSs will give you a metric of fit relative to the rest of your fits, i.e., if resid_SSs for one or more of your curves are an order of magnitude higher than the rest, you should feel suspect. Still, this is not terribly informative and is why the diagnostic_AQ_plot_function.R script exists. The `diagnostic_AQ_plot()` function will plot all of your fit curves on top of the original gas exchange data points. 

`diagnostic_AQ_plot()` has eight arguments:   

1. `curve_data = myGasExchange_data.frame`   
      + the data_frame of gas exchange data originally passed to fit_AQ_curve   
2. `fit_data = my_fit_AQ_curve_output`   
      + the data_frame of fits output by fit_AQ_curve   
3. `Photo = "myPhotoVals"`   
      + the name of the net assimilation column in the curve_data data_frame   
4. `PARi = "myPARiVals"`   
      + the name of the irradiance/PARi column in the curve_data data_frame   
5. `group_id = "myCurveIDs"`   
      + the name of the curve identifier column in the curve_data data_frame  

And, optionally:   

6. `save_to_pdf = FALSE` (or TRUE)   
      + a logical, default is FALSE. If save_to_pdf = TRUE, you must also    provide the seventh and eighth arguments. Together they will output all plots to a PDF file in the location of your choosing, with the name of your choosing.   
7. `save_path = "/Users/me/R/LightResponseCurves/Fit_Plots"`   
      + an existing file-path where you want the PDF output saved   
8. `file_name = "my_AQ_curve_fit_plots.pdf"`   
      + a name for the PDF file   

If you have only of couple of curves to interogate then the first five arguments are sufficient and you can just arrow through the plots to check out their fits. Note that the title on each plot will auto label with its group_id value so you can differentiate them. For example, in the dwarf sunflower data which we'll plot below, the group_id's are High_A, High_B, Low_A, and Low_B.
```{r load_plotting_function, echo = FALSE}
diagnostic_AQ_plot <- function(curve_data, fit_data, Photo, PARi, group_id,
                               save_to_pdf = FALSE, save_path, file_name){
      if(save_to_pdf == FALSE){ 
            par(mar = c(3, 3, 1, 1), oma = c(1, 1, 1, 1))
            for(i in seq_along(1:length(unique(curve_data[[group_id]])))){
                  single_curve <- 
                        curve_data[curve_data[[group_id]] == 
                                         unique(curve_data[[group_id]])[i],]
                  plot(
                        single_curve[[Photo]] ~ single_curve[[PARi]] ,
                        xlim = c(-2, max(curve_data[[PARi]])), 
                        ylim = c(min(curve_data[[Photo]]) - 2,
                                 max(curve_data[[Photo]]) + 2),
                        pch = 3,
                        cex = 2,
                        xlab = "",
                        ylab = "",
                        main = paste("Data from curve ",
                                     as.character(
                                           unique(single_curve[[group_id]])))
                  )
                  mtext(expression("Photo (µmol "*CO[2]*" "*m^-2*" "*s^-1*")"),
                        line = 2.4, side = 2)
                  mtext(expression("PARi (µmol photons "*m^-2*" "*s^-1*")"),
                        line = 2.4, side = 1)
                  par(new = TRUE)
                  curve(((
                        fit_data$Phi[i] * PARi + fit_data$Asat[i] - 
                              sqrt((fit_data$Phi[i] * PARi + fit_data$Asat[i])^2 - 4 *
                                         fit_data$Phi[i] * fit_data$theta[i] * PARi *
                                         fit_data$Asat[i])
                  ) / (2*fit_data$theta[i]) - fit_data$Rd[i]),
                  from = 0, to = 1600, 
                  xname = "PARi",
                  xlab = "", ylab = "", 
                  xlim = c(-2, max(curve_data[[PARi]])), 
                  ylim = c(min(curve_data[[Photo]]) - 2,
                           max(curve_data[[Photo]]) + 2),
                  axes = FALSE,
                  col = "red",
                  lwd = 2
                  )
            }} else{
             if(dir.exists(save_path)){
      pdf(paste0(save_path, file_name, ".pdf"))
      par(mar = c(3, 3, 1, 1), oma = c(1, 1, 1, 1))
      for(i in seq_along(1:length(unique(curve_data[[group_id]])))){
            single_curve <- 
                  curve_data[curve_data[[group_id]] == 
                                   unique(curve_data[[group_id]])[i],]
            plot(
                  single_curve[[Photo]] ~ single_curve[[PARi]] ,
                  xlim = c(-2, max(curve_data[[PARi]])), 
                  ylim = c(min(curve_data[[Photo]]) - 2,
                           max(curve_data[[Photo]]) + 2),
                  pch = 3,
                  cex = 2,
                  xlab = "",
                  ylab = "",
                  main = paste("Data from curve ",
                               as.character(
                                     unique(single_curve[[group_id]])))
            )
            mtext(expression("Photo (µmol "*CO[2]*" "*m^-2*" "*s^-1*")"),
                  line = 2.4, side = 2)
            mtext(expression("PARi (µmol photons "*m^-2*" "*s^-1*")"),
                  line = 2.4, side = 1)
            par(new = TRUE)
            curve(((
                  fit_data$Phi[i] * PARi + fit_data$Asat[i] - 
                        sqrt((fit_data$Phi[i] * PARi + fit_data$Asat[i])^2 - 4 *
                                   fit_data$Phi[i] * fit_data$theta[i] * PARi *
                                   fit_data$Asat[i])
            ) / (2*fit_data$theta[i]) - fit_data$Rd[i]),
            from = 0, to = 1600, 
            xname = "PARi",
            xlab = "", ylab = "", 
            xlim = c(-2, max(curve_data[[PARi]])), 
            ylim = c(min(curve_data[[Photo]]) - 2,
                     max(curve_data[[Photo]]) + 2),
            axes = FALSE,
            col = "red",
            lwd = 2
            )
            }
      dev.off()
             } else {
            return(
                  "Warning: the file path provided to save_path does not exist"
            )}
            
}
}
```

```{r prelim_AQ_plots}
diagnostic_AQ_plot(dwarf, 
                   dwarf_fits,
                   Photo = "Photo",
                   PARi = "PARi",
                   group_id = "ids")
```

As you can see this will allow you to quickly assess how well each of the fit-curves actually fits your data. These particular fits look quite good: the fit-curve falls on top of most of the original data points.

If you have more than a couple of curves, arrowing through them in the plot window is tiresome. You might want to put the plots someplace where you can more easily scroll through them or where you can save them for future reference. That is why the optional arguments exist. You can try this out with the code chunk below by removing the hashes and entering a real file path to the `save_path =` argument.

```{r saving_AQ_plots}
# diagnostic_AQ_plot(a_thal, 
#                    athal_fits,
#                    Photo = "Photo",
#                    PARi = "PARi",
#                    group_id = "ids",
#                    save_to_pdf = TRUE,
#                    save_path = "/Users/path/to/where/I/want/these",
#                    file_name = "A_thal_AQ_plots.PDF")
```
